|

### 2024-04-15 paper/paper.Rmd
update paper

### Approximated distance as an Model Violations Index (MVI)

@@

Note that since the Kullback-Leibler divergence could be affected by the number of random variables involved in the evaluation. In general, more observations will result in greater distance $D$. This does not mean $\hat{D}$ can not correctly represent the extent of the model violations since more observations in a residual plot usually lead to stronger visual signal strength as the underlying visual patterns will have higher chance to be revealed except for cases like serious overlapping. For the same data generating process, $D$ usually grows logarithmically as the number of observations increases. This is because $D = \text{log}(1 + D_{KL})$, and $D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}$ under the independence assumption.

@@

### Null distribution of the approximated distance

@@

To solve this issue while aligning with the principle of visual inference, $\hat{D}$ can be viewed as a test statistic. And the null distribution of this statistic can be approximated by the empirical distribution of $\hat{D}_{null}$, where $\hat{D}_{null}$ is the approximated distance for a null plot simulated from the fitted regression model. The approximation of the distribution involves applying the residual rotation technique [@buja2009statistical] on the fitted regression model to obtain null residuals. The null residuals are then used to make null plots and fed into the computer vision model to get predictions. The empirical distribution is constructed with the approximated distance.

There are two types of error associated with the approximation of the null distribution. These include the sampling error and the estimation error for model parameters. Increasing the number of null plots within the distribution approximation can effectively reduce sampling error. However, the estimation error for model parameters remains irreducible. The residual rotation technique operates under the assumption of a correct fitted model, using the estimated variance instead of the unknown true variance to rescale rotated residuals. Consequently, under $H_0$, the true distribution of $\hat{D}_{null}$ will slightly deviate from the approximated null distribution.

### Estimation of quantiles of the null distribution

Let $\hat{D}_{null}^{(i)}$ be the approximated distance of the $i$-th null plots, where $i = 1,...,n_{null}$ and $n_{null} \in \mathbb{N}^+$ is a sufficiently large number. Quantiles of the null distribution can then be estimated using the sample quantiles available in statistical software such as R. The details of the sample quantile computation can be found in @hyndman1996sample. In statistical testing, analysts often care about certain quantiles of the null distribution, such as the 90% quantile, the 95% quantile and the 99% quantile. These quantiles are used as critical values to decide if $H_0$ needs to be rejected. For example, if $\hat{D}$ is greater than and equal to the 95% sample quantile $Q_{null}(0.95)$, we could say the approximated distance for the actual residual plot is significantly different from the approximated distance for null plots with 95% significance level. Based on our experience, in order to get a stable estimate of the 95% quantile, $n_{null}$ usually needs to be at least $100$. And if the null distribution has a long tail, more null plots will be needed. Alternatively, a p-value can be used to represents the probability of observing an event equally or more extreme than the given event under $H_0$, and it can be estimated by $1/m\sum_{i=1}^{n_{null}}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)$.

@@

All optimized models chose to retain the additional inputs, contributing to the reduction of validation error. The number of units required for the fully-connected layer was $256$, a relatively modest number compared to the VGG16 classifier, suggesting that the problem at hand was relatively straightforward. The optimized learning rates were typically smaller than the default value recommended by Keras, which is 0.001.

@@

## Model performance

@@

The regression tree revealed that most issues arose from non-linearity problems and the presence of a second predictor in the regression model. When the error distribution had a very small variance, all computer vision models tended to under-predict the distance. Conversely, when the error distribution had a large variance, all models tended to over-predict the distance. Therefore, Additionally, for input images representing null plots, it was expected that the models will over-predict the distance, as explained in Section \ref{null-distribution-of-the-approximated-distance}.

Since most of the deviation stemmed from the presence of non-linearity violations, to further investigate this, we split the test set and re-evaluated the performance, as detailed in Table \ref{tab:performance-sub}. It was evident that metrics for null plots were notably worse compared to other categories. Furthermore, residual plots solely exhibiting non-normality issues were the easiest to predict, with very low test root mean square error (RMSE) around $0.3$, very low mean absolute error (MAE) around $0.2$, and very high $R^2$ around $0.94$. Residual plots with non-linearity issues were more challenging to assess than those with heteroskedasticity or non-normality issues. Assessing residual plots with heteroskedasticity was not as difficult as assessing those with non-linearity issues. When multiple violations were introduced to a residual plot, the performance metrics typically lay between the metrics for each individual violation.

Based on the model performance metrics, we will only use the best-performing model evaluated on the test set, namely the $32 \times 32$ model, for the subsequent analysis.

@@

train_summary %>%
  bind_rows(test_summary) %>%
  select(res, RMSE, R2, MAE, HUBER) %>%
  mutate(res = glue::glue("${res} \\times {res}$")) %>%
  mutate(across(RMSE:MAE, ~format(.x, digits = 3))) %>%
  mutate(across(HUBER, ~format(.x, digits = 2))) %>%
  kableExtra::kable(format = "latex", 
                    booktabs = TRUE, 
                    escape = FALSE,
                    col.names = c("", "RMSE", "$R^2$", "MAE", "Huber loss"),
                    caption = "The training and test performance of three optimized models with different input sizes. The best metrics are colored in red.") %>%
  kableExtra::pack_rows("Training set", 1, 3) %>%
  kableExtra::pack_rows("Test set", 4, 6) %>%
  kableExtra::column_spec(2:5, color = c("black", "red", "black",
                                       "red", "black", "black"))

@@

 # 1) root 8000 3466.12800  0.04808905  
 #   2) e_sigma>=4.108756 3288 1146.50300 -0.09147268  
 #     4) include_non_normal=FALSE 1983  817.30850 -0.22151770  
 #       8) include_heter=FALSE 1210  416.16970 -0.39866830  
 #        16) include_z=FALSE 49   20.28332 -1.22777500 * (only non-normal and large error sigma -> large over-prediction)
 #        17) include_z=TRUE 1161  360.78130 -0.36367590 *
 #       9) include_heter=TRUE 773  303.72640  0.05578156 *
 #     5) include_non_normal=TRUE 1305  244.69960  0.10613590 *
 #   3) e_sigma< 4.108756 4712 2210.89500  0.14547420  
 #     6) include_z=FALSE 2030  568.99490 -0.04100357 *
 #     7) include_z=TRUE 2682 1517.87900  0.28661890  
 #      14) include_x2=FALSE 1432  377.35510  0.07228115 *
 #      15) include_x2=TRUE 1250  999.37090  0.53216420  
 #        30) e_sigma>=0.1149429 1202  829.80300  0.49933640 *
 #        31) e_sigma< 0.1149429 48  135.83470  1.35422700  (non-linearity, second predictor, small error sigma)
 #          62) measure_striped< 0.44 40   54.18083  0.92957180 *
 #          63) measure_striped>=0.44 8   38.37423  3.47750200 * (high discreteness -> large under-prediction)

# n= 8000 
# 
# node), split, n, deviance, yval
#       * denotes terminal node
# 
#     1) root 8000 3466.1280000  0.04808905  
#       2) e_sigma>=4.108756 3288 1146.5030000 -0.09147268  
#         4) include_non_normal=FALSE 1983  817.3085000 -0.22151770  
#           8) include_heter=FALSE 1210  416.1697000 -0.39866830  
#            16) include_z=FALSE 49   20.2833200 -1.22777500  
#              32) measure_sparse< 0.3630358 7    8.9396360 -2.12605300 * (large e_sigma, null plots, not very sparse -> large over-prediction)
#              33) measure_sparse>=0.3630358 42    4.7539680 -1.07806200 * (large e_sigma, null plots, sparse -> large over-prediction)
#            17) include_z=TRUE 1161  360.7813000 -0.36367590  
#              34) measure_splines>=0.02606246 376  134.3999000 -0.59957120  
#                68) n< 143.5 116   39.6936700 -0.84719690  
#                 136) measure_monotonic>=0.07456592 7    1.3762680 -1.55893700 * (non-linearity only, small n, large e_sigma -> large over-prediction)
#                 137) measure_monotonic< 0.07456592 109   34.5436600 -0.80148880 *
#                69) n>=143.5 260   84.4197800 -0.48909200  
#                 138) e_sigma>=7.367384 63   15.0193100 -0.79693070 *
#                 139) e_sigma< 7.367384 197   61.5210600 -0.39064610  
#                   278) include_x2=FALSE 90   19.5837000 -0.53918530 *
#                   279) include_x2=TRUE 107   38.2813500 -0.26570670  
#                     558) measure_sparse< 0.5586777 84   27.3861700 -0.37645850 *
#                     559) measure_sparse>=0.5586777 23    6.1018500  0.13877850 *
#              35) measure_splines< 0.02606246 785  195.4365000 -0.25068660  
#                70) e_sigma>=5.64113 529  100.6972000 -0.37916600  
#                 140) n< 109 41   11.8877000 -0.73213180 *
#                 141) n>=109 488   83.2723700 -0.34951110  
#                   282) e_sigma>=7.90547 154   26.5350800 -0.48055150 *
#                   283) e_sigma< 7.90547 334   52.8735800 -0.28909130 *
#                71) e_sigma< 5.64113 256   67.9628700  0.01480427  
#                 142) include_x2=FALSE 113   23.8143300 -0.15341150 *
#                 143) include_x2=TRUE 143   38.4243300  0.14773000  
#                   286) n< 72.5 7    2.6335630 -0.69382120 *
#                   287) n>=72.5 136   30.5781400  0.19104510 *
#           9) include_heter=TRUE 773  303.7264000  0.05578156  
#            18) b< 32.23487 487  221.2986000 -0.09202344  
#              36) b< 0.2107948 67   26.8944500 -0.51926710  
#                72) n< 126 7    3.3765570 -1.24628700 * (large e_sigma, heteroskedasticity, small b, small n -> over-prediction)
#                73) n>=126 60   19.3863300 -0.43444810 *
#              37) b>=0.2107948 420  180.2232000 -0.02386791  
#                74) measure_splines>=0.05482434 223   77.7820400 -0.18200180  
#                 148) b< 7.647324 119   51.8788300 -0.32137070  
#                   296) measure_sparse>=0.4117995 111   43.0150500 -0.38433220 *
#                   297) measure_sparse< 0.4117995 8    2.3184780  0.55221930 *
#                 149) b>=7.647324 104   20.9469900 -0.02253160 *
#                75) measure_splines< 0.05482434 197   90.5523700  0.15513650  
#                 150) b>=5.165768 63   26.5340300 -0.02749363  
#                   300) include_x2=FALSE 35    7.9867080 -0.19247040 *
#                   301) include_x2=TRUE 28   16.4039600  0.17872730  
#                     602) measure_monotonic< 0.03127537 21    4.3016860 -0.07172379 *
#                     603) measure_monotonic>=0.03127537 7    6.8333090  0.93008060 *
#                 151) b< 5.165768 134   60.9291300  0.24099990  
#                   302) x2_sigma< 0.5865284 126   52.7919200  0.19154170  
#                     604) x2_k>=5.5 106   43.2640200  0.11519510  
#                      1208) j< 17.5 93   38.5219100  0.05750563  
#                        2416) e_even=TRUE 41   15.2536200 -0.17697420  
#                          4832) measure_sparse>=0.5392181 29    8.8138120 -0.40043160 *
#                          4833) measure_sparse< 0.5392181 12    1.4922590  0.36304760 *
#                        2417) e_even=FALSE 52   19.2367200  0.24238400  
#                          4834) measure_splines>=0.01044483 31    7.8888290 -0.02089042 *
#                          4835) measure_splines< 0.01044483 21    6.0272570  0.63102720 *
#                      1209) j>=17.5 13    2.2184070  0.52789690 *
#                     605) x2_k< 5.5 20    5.6354220  0.59617870 *
#                   303) x2_sigma>=0.5865284 8    2.9747040  1.01996600 * (large e_sigma, small spline dependence, heteroskedasticity, large x2_sigma -> large under-prediction)
#            19) b>=32.23487 286   53.6723400  0.30746350  
#              38) include_x2=FALSE 139   17.0423600  0.13425010 *
#              39) include_x2=TRUE 147   28.5161400  0.47125030 *
#         5) include_non_normal=TRUE 1305  244.6996000  0.10613590  
#          10) measure_splines>=0.08262098 455   47.2275500  0.00622856 *
#          11) measure_splines< 0.08262098 850  190.4994000  0.15961570  
#            22) include_heter=FALSE 594   22.8029300  0.08838938 *
#            23) include_heter=TRUE 256  157.6908000  0.32488300  
#              46) measure_striped>=0.47 185   31.8984100  0.17734340 *
#              47) measure_striped< 0.47 71  111.2723000  0.70931710  
#                94) e_even=FALSE 32   18.0084200  0.22023240  
#                 188) a>=-0.205314 21    5.0433870 -0.06070074 *
#                 189) a< -0.205314 11    8.1435320  0.75655920 *
#                95) e_even=TRUE 39   79.3286800  1.11061700  
#                 190) b< 2.373839 11    0.2423938  0.06215229 *
#                 191) b>=2.373839 28   62.2437600  1.52251400  (discrete e_dist,  high discreteness, heteroskedasticity, small splines dependence, non-normality, large e_sigma -> large under-prediction)
#                   382) x2_dist=even_discrete,lognormal,uniform 19   33.0782400  1.01336800 *
#                   383) x2_dist=normal 9   13.8421700  2.59737800 *
#       3) e_sigma< 4.108756 4712 2210.8950000  0.14547420  
#         6) include_z=FALSE 2030  568.9949000 -0.04100357  
#          12) b< 28.58967 1159  316.0764000 -0.12131010  
#            24) include_non_normal=FALSE 441  185.6341000 -0.33971870  
#              48) include_heter=FALSE 60    8.3705760 -1.15981800 * (null plots, small e_sigma -> large over-prediction)
#              49) include_heter=TRUE 381  130.5548000 -0.21056910  
#                98) b< 0.4619795 85   22.8319800 -0.60512210 *
#                99) b>=0.4619795 296   90.6909300 -0.09726844  
#                 198) b>=1.080411 252   72.3919200 -0.15604160  
#                   396) b< 2.878015 65   20.2765800 -0.37135770 *
#                   397) b>=2.878015 187   48.0544100 -0.08119905 *
#                 199) b< 1.080411 44   12.4430700  0.23934130 *
#            25) include_non_normal=TRUE 718   96.4846700  0.01283772 *
#          13) b>=28.58967 871  235.4979000  0.06585671  
#            26) include_heter=FALSE 361  111.6977000 -0.10528680  
#              52) include_non_normal=FALSE 46    6.2078660 -1.13547200 * (null plots, small e_sigma -> large over-prediction)
#              53) include_non_normal=TRUE 315   49.5417500  0.04515286  
#               106) e_sigma>=0.2273127 247   17.7293100 -0.01172324 *
#               107) e_sigma< 0.2273127 68   28.1111100  0.25174690  
#                 214) measure_splines>=0.03710735 23   10.7097800 -0.09471551 *
#                 215) measure_splines< 0.03710735 45   13.2294000  0.42882770 *
#            27) include_heter=TRUE 510  105.7420000  0.18699950  
#              54) include_non_normal=TRUE 235   41.6282300  0.05417806 *
#              55) include_non_normal=FALSE 275   56.4252500  0.30050140  
#               110) include_x2=FALSE 137   20.9155600  0.17983570 *
#               111) include_x2=TRUE 138   31.5346400  0.42029280  
#                 222) b< 70.08912 96   17.2702500  0.30834030 *
#                 223) b>=70.08912 42   10.3110100  0.67618420 *
#         7) include_z=TRUE 2682 1517.8790000  0.28661890  
#          14) include_x2=FALSE 1432  377.3551000  0.07228115  
#            28) measure_splines>=0.07058104 731  178.2000000 -0.01483915  
#              56) e_sigma>=2.44726 138   40.9281500 -0.18500620  
#               112) include_heter=FALSE 59   25.9010700 -0.41304000  
#                 224) include_non_normal=FALSE 38   17.2724700 -0.68141390  
#                   448) measure_splines>=0.1003984 20    5.6757540 -1.07654300 * (non-linearity only, moderate e_sigma -> large over-prediction)
#                   449) measure_splines< 0.1003984 18    5.0046810 -0.24238140 *
#                 225) include_non_normal=TRUE 21    0.9391262  0.07258882 *
#               113) include_heter=TRUE 79    9.6678490 -0.01470237 *
#              57) e_sigma< 2.44726 593  132.3458000  0.02476126  
#               114) e_sigma>=0.158351 527  105.6393000 -0.00147316 *
#               115) e_sigma< 0.158351 66   23.4476700  0.23423920  
#                 230) b< 71.04535 55   15.9627800  0.12578350 *
#                 231) b>=71.04535 11    3.6032230  0.77651720 *
#            29) measure_splines< 0.07058104 701  187.8212000  0.16312990  
#              58) n< 62.5 33   14.3260500 -0.31035730 *
#              59) n>=62.5 668  165.7314000  0.18652070 *
#          15) include_x2=TRUE 1250  999.3709000  0.53216420  
#            30) e_sigma>=0.1149429 1202  829.8030000  0.49933640  
#              60) e_sigma>=2.906419 328  129.1838000  0.27288400  
#               120) measure_splines>=0.04413018 134   50.5691600  0.09851024  
#                 240) b< 2.910823 23   10.1825400 -0.20213530  
#                   480) x2_k>=6.5 15    4.3945260 -0.49662390 *
#                   481) x2_k< 6.5 8    2.0480610  0.35003080 *
#                 241) b>=2.910823 111   37.8769400  0.16080620  
#                   482) b>=7.320158 97   22.8753700  0.08661750 *
#                   483) b< 7.320158 14   10.7686400  0.67482760 *
#               121) measure_splines< 0.04413018 194   71.7259100  0.39332780  
#                 242) j< 7.5 60   32.0807400  0.19368700  
#                   484) x2_sigma< 0.343438 10   12.3504400 -0.51629950 *
#                   485) x2_sigma>=0.343438 50   13.6813400  0.33568430 *
#                 243) j>=7.5 134   36.1830100  0.48271920 *
#              61) e_sigma< 2.906419 874  677.4868000  0.58432080  
#               122) measure_splines>=0.04519141 613  424.1886000  0.45960850  
#                 244) n>=202.5 393  209.4163000  0.32432050  
#                   488) include_non_normal=TRUE 188   90.6250100  0.18780810  
#                     976) x2_k< 7.5 101   41.5156400  0.04431936 *
#                     977) x2_k>=7.5 87   44.6157600  0.35438700  
#                      1954) measure_sparse>=0.391553 80   35.8617500  0.29057300 *
#                      1955) measure_sparse< 0.391553 7    4.7050420  1.08369000 * (not sparse, non-normality, large n, small e_sigma, second predictor, non-linearity -> large under-prediction)
#                   489) include_non_normal=FALSE 205  112.0748000  0.44951240  
#                     978) e_sigma>=0.2441489 175   86.5106600  0.36295000  
#                      1956) measure_splines>=0.07729252 127   55.6660800  0.25807550 *
#                      1957) measure_splines< 0.07729252 48   25.7519700  0.64043040  
#                        3914) b>=21.95702 24    7.5694370  0.34280130 *
#                        3915) b< 21.95702 24   13.9305500  0.93805950  
#                          7830) x2_dist=even_discrete,uniform 13    2.6412500  0.51209270 *
#                          7831) x2_dist=lognormal,normal 11    6.1427810  1.44147500 *
#                     979) e_sigma< 0.2441489 30   16.6036900  0.95445990  
#                      1958) a< -0.6962934 8    1.2723410  0.16916930 *
#                      1959) a>=-0.6962934 22    8.6039200  1.24002000 *
#                 245) n< 202.5 220  194.7300000  0.70128210  
#                   490) measure_sparse>=0.6256282 81   38.6326300  0.36079080  
#                     980) measure_monotonic>=0.237124 8    0.2987567 -0.26782100 *
#                     981) measure_monotonic< 0.237124 73   34.8262100  0.42967980 *
#                   491) measure_sparse< 0.6256282 139  141.2344000  0.89969790  
#                     982) n>=97 88   72.2102700  0.72131320  
#                      1964) x1_sigma>=0.4241299 54   26.0940300  0.51943470  
#                        3928) e_sigma>=0.2485052 45   19.5310800  0.39828410  
#                          7856) x1_sigma< 0.5588507 38   11.8880800  0.26476900 *
#                          7857) x1_sigma>=0.5588507 7    3.2882920  1.12308000 *
#                        3929) e_sigma< 0.2485052 9    2.6000360  1.12518800 *
#                      1965) x1_sigma< 0.4241299 34   40.4201400  1.04194400  
#                        3930) j< 7.5 11    8.6545230  0.32377180 *
#                        3931) j>=7.5 23   23.3787300  1.38541700  
#                          7862) b< 20.93247 14   10.8261000  0.95400130 *
#                          7863) b>=20.93247 9    5.8936770  2.05650900 * (moderate b, large j, small x1_sigma, small n, small sparse, second predictor, non-linearity -> large under-prediction)
#                     983) n< 97 51   61.3921000  1.20749900  
#                      1966) e_dist=uniform 16    6.5588030  0.49547860 *
#                      1967) e_dist=even_discrete,lognormal 35   43.0135900  1.53299400  
#                        3934) measure_monotonic>=0.04508442 17   19.7133200  0.99850550 *
#                        3935) measure_monotonic< 0.04508442 18   13.8570300  2.03778800 *
#               123) measure_splines< 0.04519141 261  221.3718000  0.87722750  
#                 246) measure_sparse>=0.654522 53   26.6382500  0.36906170  
#                   492) measure_striped>=0.57 11    2.8015390 -0.20941000 *
#                   493) measure_striped< 0.57 42   19.1917400  0.52056620 *
#                 247) measure_sparse< 0.654522 208  177.5598000  1.00671200  
#                   494) e_sigma>=2.064604 95   37.3057400  0.73978960  
#                     988) n< 94.5 10    2.7534790  0.01814227 *
#                     989) n>=94.5 85   28.7318400  0.82468930  
#                      1978) measure_sparse< 0.3468051 11    2.5461540  0.27928850 *
#                      1979) measure_sparse>=0.3468051 74   22.4272100  0.90576240 *
#                   495) e_sigma< 2.064604 113  127.7952000  1.23111600  
#                     990) measure_splines>=0.03040418 40   30.6303500  0.78917000  
#                      1980) include_non_normal=TRUE 16    4.6253120  0.29159560 *
#                      1981) include_non_normal=FALSE 24   19.4029000  1.12088600 *
#                     991) measure_splines< 0.03040418 73   85.0712900  1.47327800  
#                      1982) b< 90.99624 66   59.3920200  1.35710600  
#                        3964) a>=-0.6695046 52   41.5200500  1.19796400  
#                          7928) x2_sigma< 0.4287885 20   11.8256700  0.80120150  
#                           15856) measure_splines>=0.009178431 7    2.5265790  0.21136300 *
#                           15857) measure_splines< 0.009178431 13    5.5523770  1.11880700 *
#                          7929) x2_sigma>=0.4287885 32   24.5782200  1.44594000  
#                           15858) x2_sigma>=0.4851646 24   12.4345200  1.18424600  
#                             31716) e_dist=uniform 7    0.5259639  0.41482310 *
#                             31717) e_dist=even_discrete,lognormal 17    6.0580960  1.50106700 *
#                           15859) x2_sigma< 0.4851646 8    5.5692240  2.23102400 * (small x2_sigma, small b, small splines dependence, small e_sigma, small sparse, second predictor, non-linearity -> large under-prediction)
#                        3965) a< -0.6695046 14   11.6634200  1.94820600 *
#                      1983) b>=90.99624 7   16.3901800  2.56861600 * (large b, small splines dependence, small e_sigma, small sparse, second predictor, non-linearity -> large under-prediction)
#            31) e_sigma< 0.1149429 48  135.8347000  1.35422700  
#              62) measure_striped< 0.44 40   54.1808300  0.92957180  
#               124) include_non_normal=TRUE 21   14.0872600  0.39910760  
#                 248) include_heter=TRUE 13    5.7917480  0.06683663 *
#                 249) include_heter=FALSE 8    4.5279800  0.93904800 *
#               125) include_non_normal=FALSE 19   27.6530600  1.51587400 *
#              63) measure_striped>=0.44 8   38.3742300  3.47750200 * (high discreteness, small e_sigma, second predictor, non-linearity -> large under-prediction)

model_pred %>%
  filter(res == 32L) %>%
  filter(type == "test") %>%
  mutate(.resid = effect_size - vss) %>%
  mutate(include_z = factor(include_z),
         include_heter = factor(include_heter),
         include_x2 = factor(include_x2),
         include_non_normal = factor(include_non_normal),
         e_even = factor(e_even),
         x1_even = factor(x1_even),
         x2_even = factor(x2_even)) %>%
  select(-filename, -vss, -res, -plot_uid, -data_type, -effect_size, -type) %>%
  rpart::rpart(.resid ~ ., data = ., control = rpart::rpart.control(cp = 0.001), model = TRUE) -> test_tree
```

@@

model_pred %>%
  filter(res == 32L) %>%
  group_by(type, include_non_normal, include_heter, include_z) %>%
  summarise(n = n(), 
            RMSE = yardstick::rmse_vec(effect_size, vss),
            R2 = yardstick::rsq_vec(effect_size, vss),
            MAE = yardstick::mae_vec(effect_size, vss),
            HUBER = yardstick::huber_loss_vec(effect_size, vss)) %>%
  mutate(violations = ifelse(include_z, "non-linearity", "null")) %>%
  mutate(violations = ifelse(include_heter, glue("{violations} + heteroskedasticity"), violations)) %>%
  mutate(violations = ifelse(include_non_normal, glue("{violations} + non-normality"), violations)) %>%
  mutate(violations = gsub("null \\+ ", "", violations)) %>%
  ungroup() %>%
  select(type, violations, n, RMSE, R2, MAE, HUBER) %>%
  arrange(desc(type)) %>%
  select(-type) %>%
  mutate(n = format(n)) %>%
  mutate(across(RMSE:MAE, ~format(.x, digits = 3))) %>%
  mutate(across(HUBER, ~format(.x, digits = 2))) %>%
  mutate(violations = ifelse(violations == "null", "no violations", violations)) %>%
  mutate(R2 = gsub(".*NA.*", "", R2)) %>%
  kableExtra::kable("latex", 
                    booktabs = TRUE,
                    escape = FALSE,
                    col.names = c("Violations", "\\#samples", "RMSE", "$R^2$", "MAE", "Huber loss"),
                    caption = "The training and test performance of the $32 \\times 32$ model presented with different model violations.  The best metrics are colored in red.") %>%
  kableExtra::kable_styling(latex_options = c("scale_down")) %>%
  kableExtra::pack_rows("Training set", 1, 8) %>%
  kableExtra::pack_rows("Test set", 9, 16) %>%
  kableExtra::column_spec(3, color = c(rep("black", 4), "red", rep("black", 3),
                                       rep("black", 4), "red", rep("black", 3))) %>%
  kableExtra::column_spec(4, color = c(rep("black", 4), "red", rep("black", 3),
                                       rep("black", 4), "red", rep("black", 3))) %>%
  kableExtra::column_spec(5, color = c(rep("black", 4), "red", rep("black", 3),
                                       rep("black", 4), "red", rep("black", 3))) %>%
  kableExtra::column_spec(6, color = c(rep("black", 4), "red", rep("black", 3),
                                       rep("black", 4), "red", rep("black", 3)))
```

## Model Violations Index (MVI)

Figures \ref{fig:poly-index} and \ref{fig:heter-index} display the residual plots for fitted models exhibiting varying degrees of non-linearity and heteroskedasticity. Each residual plot's Model Violation Index (MVI) is computed using Equation \ref{eq:mvi} with $C = 10$. When $\text{MVI} > 8$, the visual patterns are notably strong and easily discernible by humans. In the range $6 < \text{MVI} < 8$, the visibility of the visual pattern diminishes as MVI decreases. Conversely, when $\text{MVI} < 6$, the visual pattern tends to become relatively faint and challenging to observe. Table \ref{tab:mvi} provides a summary of the MVI usage and it is applicable to other linear regression models.

@@

For each lineup used in @li2023plot, there is one actual residual plot and 19 null plots. While the distance $D$ for the actual residual plot depends on the underlying data generating process, the distance $D$ for the null plots is zero. We have used our optimized computer vision model to predict $\hat{D}$ for both the actual residual plots and the null plots. Additionally, the appropriate conventional tests including the Ramsey Regression Equation Specification Error Test (RESET) [@ramsey1969tests] for non-linearity and the Breusch-Pagan test  [@breusch1979simple] for heteroskedasticity were applied on the same data for comparison.

@@

>>>>

### Model Violations Index (MVI) {#sec:model-violations-index}

@@

@@

### Lineup evaluation

@@

To address this issue while adhering to the principle of visual inference, we can compare the estimated distance $\hat{D}$ to the approximated distances for the null plots in a lineup. Specifically, if a lineup comprises 20 plots, the null hypothesis $H_0$ will be rejected if $\hat{D}$ exceeds the maximum approximated distance among the $m - 1$ null plots, denoted as $\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}$, where $\hat{D}_{null}^{(i)}$ represents the approximated distance for the $i$-th null plot. This approach is equivalent to the typical lineup protocol requiring a 95% significance level, where $H_0$ is rejected if the data plot is identified as the most distinct plot by the sole participant. The approximated distance serves as a metric to quantify the difference between the data plot and the null plots, as intended.

For lineups consisting of more than 20 plots, the 95% significance level can be maintained if the number of plots is a multiple of 20. Specifically, for lineups comprising $20p$ plots, where $p$ is a positive integer, we reject $H_0$ if $\hat{D}$ exceeds 95% ${\hat{D}_{null}^{(i)}}$ for $i = 1, \ldots, 20p-1$. The p-value in this case is given by $\frac{1}{m} + \frac{1}{m}\sum{i=1}^{m-1} I\left(\hat{D}_{null}^{(i)} > \hat{D}\right)$, where $I(\cdot)$ is the indicator function.

Moreover, if the number of plots in a lineup, denoted by $m$, is sufficiently large, the empirical distribution of ${\hat{D}_{null}^{(i)}}$ can be viewed as an approximation of the null distribution of the approximated distance. Consequently, quantiles of the null distribution can be estimated using the sample quantiles available in statistical software such as R, and these quantiles can be utilized for decision-making purposes. The details of the sample quantile computation can be found in @hyndman1996sample. For instance, if $\hat{D}$ is greater than or equal to the 95% sample quantile, denoted as $Q_{null}(0.95)$, we can conclude that the approximated distance for the actual residual plot is significantly different from the approximated distance for null plots with a 95% significance level. Based on our experience, to obtain a stable estimate of the 95% quantile, the number of null plots, $n_{null}$, typically needs to be at least 100. However, if the null distribution exhibits a long tail, a larger number of null plots may be required. Alternatively, a p-value can be used to represent the probability of observing an event equally or more extreme than the given event under the null hypothesis $H_0$, and it can be estimated by $\frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)$.

@@

All optimized models chose to retain the additional inputs, contributing to the reduction of validation error. The number of units required for the fully-connected layer was $256$, a relatively modest number compared to the VGG16 classifier, suggesting that the problem at hand was less complex. The optimized learning rates were smaller than the default value recommended by Keras, which is 0.001.

@@

## Summary of model performance

@@

The regression tree revealed that most issues arose from non-linearity problems and the presence of a second predictor in the regression model. When the error distribution had a very small variance, all computer vision models tended to under-predict the distance. Conversely, when the error distribution had a large variance, all models tended to over-predict the distance. Furthermore, for input images representing null plots, it was expected that the models will over-predict the distance, as explained in Section \ref{lineup-evaluation}.

Since most of the deviation stemmed from the presence of non-linearity violations, to further investigate this, we split the test set and re-evaluated the performance, as detailed in Table \ref{tab:performance-sub}. It was evident that metrics for null plots were notably worse compared to other categories. Furthermore, residual plots solely exhibiting non-normality issues were the easiest to predict, with very low test root mean square error (RMSE) at around $0.3$. Residual plots with non-linearity issues were more challenging to assess than those with heteroskedasticity or non-normality issues. Assessing residual plots with heteroskedasticity was not as difficult as assessing those with non-linearity issues. When multiple violations were introduced to a residual plot, the performance metrics typically lay between the metrics for each individual violation.

Based on the model performance metrics, we chose to use the best-performing model evaluated on the test set, namely the $32 \times 32$ model, for the subsequent analysis.

@@

train_summary %>%
  bind_rows(test_summary) %>%
  select(res, RMSE, R2, MAE, HUBER) %>%
  mutate(res = glue::glue("${res} \\times {res}$")) %>%
  mutate(across(RMSE:MAE, ~format(.x, digits = 3))) %>%
  mutate(across(HUBER, ~format(.x, digits = 2))) %>%
  kableExtra::kable(format = "latex", 
                    booktabs = TRUE, 
                    escape = FALSE,
                    col.names = c("", "RMSE", "$R^2$", "MAE", "Huber loss"),
                    caption = "The training and test performance of three optimized models with different input sizes.") %>%
  kableExtra::pack_rows("Training set", 1, 3) %>%
  kableExtra::pack_rows("Test set", 4, 6)
```

@@

@@

model_pred %>%
  filter(res == 32L) %>%
  group_by(include_non_normal, include_heter, include_z) %>%
  summarise(train_n = sum(type == "train"),
            train_RMSE = yardstick::rmse_vec(effect_size[type == "train"], vss[type == "train"]),
            test_n = sum(type == "test"),
            test_RMSE = yardstick::rmse_vec(effect_size[type == "test"], vss[type == "test"])) %>%
  mutate(violations = ifelse(include_z, "non-linearity", "null")) %>%
  mutate(violations = ifelse(include_heter, glue("{violations} + heteroskedasticity"), violations)) %>%
  mutate(violations = ifelse(include_non_normal, glue("{violations} + non-normality"), violations)) %>%
  mutate(violations = gsub("null \\+ ", "", violations)) %>%
  ungroup() %>%
  select(violations, train_n, train_RMSE, test_n, test_RMSE) %>%
  mutate(train_n = format(train_n)) %>%
  mutate(test_n = format(test_n)) %>%
  mutate(across(c(train_RMSE, test_RMSE), ~format(.x, digits = 3))) %>%
  mutate(violations = ifelse(violations == "null", "no violations", violations)) %>%
  kableExtra::kable("latex", 
                    booktabs = TRUE,
                    escape = FALSE,
                    col.names = c("Violations", "\\#training samples", "Training RMSE", "\\#test samples", "Test RMSE"),
                    caption = "The training and test performance of the $32 \\times 32$ model presented with different model violations.") %>%
  kableExtra::kable_styling(latex_options = c("scale_down"))
```

## Demostration of MVI

As described in Section \ref{sec:model-violations-index} a Model Violations Index (MVI) can be constructed using $\hat{D}$. This section provides guidelines and examples of MVI.

Figures \ref{fig:poly-index} and \ref{fig:heter-index} display the residual plots for fitted models exhibiting varying degrees of non-linearity and heteroskedasticity. Each residual plot's MVI is computed using Equation \ref{eq:mvi} with $C = 10$. When $\text{MVI} > 8$, the visual patterns are notably strong and easily discernible by humans. In the range $6 < \text{MVI} < 8$, the visibility of the visual pattern diminishes as MVI decreases. Conversely, when $\text{MVI} < 6$, the visual pattern tends to become relatively faint and challenging to observe. Table \ref{tab:mvi} provides a summary of the MVI usage and it is applicable to other linear regression models.

@@

For each lineup used in @li2023plot, there is one actual residual plot and 19 null plots. While the distance $D$ for the actual residual plot depends on the underlying data generating process, the distance $D$ for the null plots is zero. We have used our optimized computer vision model to predict the approximated distance for both the actual residual plots and the null plots. To have a fair comparison, $H_0$ will be rejected if the actual residual plot has the greatest approximated distance among all plots in a lineup. Additionally, the appropriate conventional tests including the Ramsey Regression Equation Specification Error Test (RESET) [@ramsey1969tests] for non-linearity and the Breusch-Pagan test  [@breusch1979simple] for heteroskedasticity were applied on the same data for comparison.

@@

```{r eval = FALSE}
plot_path <- my_vi$plot_resid() %>%
  my_vi$save_plot()
input_auxiliary <- my_vi$auxiliary()
